---
title: Effect of normalization on the PBMC 4K data set
author: Aaron Lun
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc_float: true
---    

```{r}
knitr::opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE)
set.seed(10000)    
```

# Loading in the data

Here we use the PBMC 4K data set from 10X Genomics.

```{r}
library(BiocFileCache)
bfc <- BiocFileCache(ask = FALSE)
raw.path <- bfcrpath(bfc, file.path("http://cf.10xgenomics.com/samples",
    "cell-exp/2.1.0/pbmc4k/pbmc4k_filtered_gene_bc_matrices.tar.gz"))
out.dir <- file.path(tempdir(), "pbmc4k")
untar(raw.path, exdir=out.dir)

library(DropletUtils)
fname <- file.path(out.dir, "filtered_gene_bc_matrices/GRCh38")
sce <- read10xCounts(fname, col.names=TRUE)
sce
```

# Evaluating size factors

We compute size factors with a variety of strategies:

```{r}
library(FurtherNorm2018)
all.sf <- runAllMethods(counts(sce), threshold=0.1)
lapply(all.sf, summary)
```

If the size factors are accurate, per-cluster normalization factors computed from normalized counts should be close to 1.
Otherwise, if any composition biases are still present, we should see that in the comparisons between cluster average profiles.
First, we download the clustering information:

```{r}
clust.path <- bfcrpath(bfc,
    "http://cf.10xgenomics.com/samples/cell-exp/2.1.0/pbmc4k/pbmc4k_analysis.tar.gz")
clust.dir <- file.path(tempdir(), "clustering")
untar(clust.path, exdir=clust.dir)

clusters <- read.csv(file.path(clust.dir, "analysis/clustering",
    "kmeans_10_clusters/clusters.csv"))
clusters <- clusters[match(colnames(sce), clusters$Barcode),]
head(clusters)
```

We then assess the performance of each clustering method based on the presence of composition biases between clusters.
We remove very small clusters for which calculations are likely to be inaccurate:

```{r}
cell.types <- clusters$Cluster
freq <- table(cell.types)
to.use <- cell.types %in% names(freq)[freq > 20]
lapply(all.sf, FUN=function(sf) { 
    clusterBias(counts(sce)[,to.use], sf[to.use], clust=cell.types[to.use]) 
})
```

Indeed, we can see that library size normalization exhibits a ~30% bias for one of the clusters.

# Examining downstream effects

We will focus on clusters 1 and 6, which have the most obvious difference in terms of composition bias:

```{r}
sf.l <- all.sf$Lib
sf.q <- all.sf$QuickSum
plot(sf.l, sf.q, xlab="Library size factors", ylab="Quicksum size factors", log="xy")
is.1 <- which(cell.types=="1")
points(sf.l[is.1], sf.q[is.1], col="orange")
is.6 <- which(cell.types=="6")
points(sf.l[is.6], sf.q[is.6], col="dodgerblue")
```

Testing for DE between clusters with `r Biocpkg("edgeR")`:
Clearly there is a difference when we account for composition biases:

```{r}
keep <- scater::calcAverage(sce) > 0.1
subcounts <- counts(sce)[keep,]
res.q <- testDE(subcounts, sf.q, is.1, is.6)
res.l <- testDE(subcounts, sf.l, is.1, is.6)

library(edgeR)
table(decideTestsDGE(res.q), decideTestsDGE(res.l))
```

By comparison, `computeSumFactors()` and `quickSumFactors()` are in greater agreement:
    
```{r}
sf.d <- all.sf$Deconv.clust
res.d <- testDE(subcounts, sf.d, is.1, is.6)
table(decideTestsDGE(res.d), decideTestsDGE(res.q))
```

# Session information

```{r}
sessionInfo()
```
