---
title: Effect of normalization on the workflow-processed data 
author: Aaron Lun
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc_float: true
---    

```{r, echo=FALSE}
knitr::opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE)
```

# Setup 

This document assesses the performance of the different normalization methods on the processed `SingleCellExperiment` objects from the `r Biocpkg("simpleSingleCell")` workflow.
Those workflows need to be compiled first, see this [repository](https://github.com/MarioniLab/simpleSingleCellResults) for details.

```{r}
library(FurtherNorm2018)
library(scran)
library(scater)

# Change according to the workflow location on your machine:
dir <- "~/AaronDocs/Research/simpleSingleCell/package/vignettes/"
```

# 416B

Loading:

```{r}
sce <- readRDS(file.path(dir, "416B_data.rds"))
table(sce$cluster)
```

Performing normalization without any clustering:

```{r}
sce <- computeSumFactors(sce)
clusterBias(counts(sce), sizeFactors(sce), sce$cluster, threshold=1)
```

We can also re-run it with some clustering:

```{r}
clust <- quickCluster(sce)
table(clust, sce$cluster)
sce <- computeSumFactors(sce, cluster=clust)
clusterBias(counts(sce), sizeFactors(sce), sce$cluster, threshold=1)
```

And for comparison, we use library sizes:

```{r}
clusterBias(counts(sce), librarySizeFactors(sce), sce$cluster, threshold=1)
```

```{r, echo=FALSE, results="hide"}
gc()
```

# Brain 

Loading:

```{r}
sce <- readRDS(file.path(dir, "brain_data.rds"))
table(sce$cluster)
```

Performing normalization without any clustering:

```{r}
sce <- computeSumFactors(sce, min.mean=0.1)
clusterBias(counts(sce), sizeFactors(sce), sce$cluster, threshold=0.1)
```

We can also re-run it with some clustering:

```{r}
clust <- quickCluster(sce, pc.approx=TRUE)
table(clust, sce$cluster)
sce <- computeSumFactors(sce, cluster=clust, min.mean=0.1)
clusterBias(counts(sce), sizeFactors(sce), sce$cluster, threshold=0.1)
```

And for comparison, we use library sizes:

```{r}
clusterBias(counts(sce), librarySizeFactors(sce), sce$cluster, threshold=0.1)
```

```{r, echo=FALSE, results="hide"}
gc()
```

# PBMC 

Loading:

```{r}
sce <- readRDS(file.path(dir, "pbmc_data.rds"))
table(sce$Cluster)
```

Performing normalization without any clustering:

```{r}
sce <- computeSumFactors(sce, min.mean=0.1)
clusterBias(counts(sce), sizeFactors(sce), sce$Cluster, threshold=0.1)
```

We can also re-run it with some clustering:

```{r}
clust <- quickCluster(sce, pc.approx=TRUE)
table(clust, sce$Cluster)
sce <- computeSumFactors(sce, cluster=clust, min.mean=0.1)
clusterBias(counts(sce), sizeFactors(sce), sce$Cluster, threshold=0.1)
```

And for comparison, we use library sizes:

```{r}
clusterBias(counts(sce), librarySizeFactors(sce), sce$Cluster, threshold=0.1)
```

```{r, echo=FALSE, results="hide"}
gc()
```

# Mammary gland 

Loading:

```{r}
sce <- readRDS(file.path(dir, "mammary.rds"))
table(sce$Cluster)
```

Performing normalization without any clustering:

```{r}
sce <- computeSumFactors(sce, min.mean=0.1)
clusterBias(counts(sce), sizeFactors(sce), sce$Cluster, threshold=0.1)
```

We can also re-run it with some clustering:

```{r}
clust <- quickCluster(sce, pc.approx=TRUE)
table(clust, sce$Cluster)
sce <- computeSumFactors(sce, cluster=clust, min.mean=0.1)
clusterBias(counts(sce), sizeFactors(sce), sce$Cluster, threshold=0.1)
```

And for comparison, we use library sizes:

```{r}
clusterBias(counts(sce), librarySizeFactors(sce), sce$Cluster, threshold=0.1)
```

```{r, echo=FALSE, results="hide"}
gc()
```

# Concluding remarks

Deconvolution with pre-clustering generally yields good results, with size factors close to unity for all final clusters.
Error seems to be limited to a maximum of 15% in most cases, which is acceptable for log-fold change considerations.
Without pre-clustering, maximum errors are around 20%, and can be higher still with library size normalization, e.g., in the brian dataset.

The major exception occurs in the PBMC data set where several clusters are far from 1 with pre-clustering and deconvolution.
The most extreme case is due to discreteness in `clusterBias()` itself and can be ignored.
The others are milder and are probably caused by coercion of multiple small clusters into a single level in `clust` to satisfy the minimum size in `quickCluster()`.
As a result, the non-DE majority is violated, which results in 20-30% errors (consistent with the simulations).
I consider this a pathological case as it involves several small clusters that are very different.

# Session info

```{r}
sessionInfo()
```

