---
title: Effect of normalization on the Zeisel data set
author: Aaron Lun
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc_float: true
---    

```{r}
knitr::opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE)
set.seed(100000)
```

# Loading in the data

The Zeisel data set has a large number of cells across a heterogeneous population with low UMI counts.
Thus, it is a perfect example of the type of scRNA-seq data that has composition biases _and_ many zeroes.

```{r}
readFormat <- function(infile) { 
    # First column is empty.
    metadata <- read.delim(infile, stringsAsFactors=FALSE, header=FALSE, nrow=10)[,-1] 
    rownames(metadata) <- metadata[,1]
    metadata <- metadata[,-1]
    metadata <- as.data.frame(t(metadata))

    # First column after row names is some useless filler.
    counts <- read.delim(infile, stringsAsFactors=FALSE, header=FALSE, row.names=1, skip=11)[,-1] 
    counts <- as.matrix(counts)
    return(list(metadata=metadata, counts=counts))
}

library(BiocFileCache)
bfc <- BiocFileCache(ask = FALSE)
base.url <- file.path("https://storage.googleapis.com",
    "linnarsson-lab-www-blobs/blobs/cortex")
mRNA.path <- bfcrpath(bfc, file.path(base.url, 
    "expression_mRNA_17-Aug-2014.txt"))
cellular <- readFormat(mRNA.path)
counts <- cellular$counts
```

# Evaluating size factors

We compute size factors with a variety of strategies:

```{r}
library(FurtherNorm2018)
all.sf <- runAllMethods(counts, threshold=0.1)
lapply(all.sf, summary)
```

If the size factors are accurate, per-cluster normalization factors computed from normalized counts should be close to 1.
Otherwise, if any composition biases are still present, we should see that in the comparisons between cluster average profiles.

```{r}
cell.types <- cellular$metadata$level1class
lapply(all.sf, FUN=clusterBias, counts=counts, clust=cell.types)
```

Indeed, we can see that library size normalization exhibits a ~30% bias for one of the clusters.

# Examining downstream effects

We will focus on oligodendrocytes and pyramidal neurons, which have an obvious difference in terms of composition bias:

```{r}
sf.l <- all.sf$Lib
sf.q <- all.sf$QuickSum
plot(sf.l, sf.q, xlab="Library size factors", ylab="Quicksum size factors", log="xy")
is.oligo <- which(cell.types=="oligodendrocytes")
points(sf.l[is.oligo], sf.q[is.oligo], col="orange")
is.pyramidal <- which(cell.types=="pyramidal CA1")
points(sf.l[is.pyramidal], sf.q[is.pyramidal], col="dodgerblue")
```

Testing for DE between oligodendrocytes and pyramidal neurons with `r Biocpkg("edgeR")`.
Clearly there is a difference when we account for composition biases:

```{r}
keep <- scater::calcAverage(counts) > 0.1
subcounts <- counts[keep,]
res.q <- testDE(subcounts, sf.q, which(cell.types=="oligodendrocytes"), which(cell.types=="pyramidal CA1"))
res.l <- testDE(subcounts, sf.l, which(cell.types=="oligodendrocytes"), which(cell.types=="pyramidal CA1"))

library(edgeR)
table(decideTestsDGE(res.q), decideTestsDGE(res.l))
```

By comparison, `computeSumFactors()` and `quickSumFactors()` are in greater agreement:

```{r}
sf.d <- all.sf$Deconv.clust
res.d <- testDE(subcounts, sf.d, which(cell.types=="oligodendrocytes"), which(cell.types=="pyramidal CA1"))
table(decideTestsDGE(res.d), decideTestsDGE(res.q))
```

# Session information

```{r}
sessionInfo()
```
